Edge sign prediction in general graphs and Correlation Clustering

### 2017-10-14 edge/model.tex
describe edge matrix formulation in more detail
explain Frank Wolfe better, as well as the kmean baseline and its
iterative improvement

Based on these specific definitions of $s$ and $g$, we now formally write down the problem of
finding and assigning the most informative directions to every edge of $G$
\begin{problem}[]
  \label{p:edge_full}
  Given a graph $G=(V, E)$, node profiles $X\in [-1, 1]^{n\times d}$ and an integer $k \in \Nbb$,
  find a set of $k$ unit-norm directions $\mathcal{D}_k = \{w_1, w_2, \ldots, w_k\} \in \dsphere$
  and associate to every edge of $E$ the direction with the maximal goodness. Formally, solve
  \begin{equation}
    \label{eq:edge_full}
    \argmax_{w_1,\ldots,w_k \in \dsphere}
    \sum_{u,v \in E} \max_{\ell \in \rangesk} g(s_{uv}, w_{\ell}) =
    \sum_{u,v \in E} \max_{\ell \in \rangesk} \left(x_u \circ x_v\right)^T w_{\ell} = f(k)
  \end{equation}
\end{problem}

@@

If $k=|E|$, the topology of the graph becomes irrelevant, as we can simply exploit the trivial
solution of having a single direction $w_{uv} = \frac{s_{uv}}{||s_{uv}||}$ for every edge $(u,v)$.
More generally, as $k$ increases, we expect the value $\max_{\mathcal{D}_k} f(k)$ to increases as
well, at the cost of interpretability. There could be a principled approach to finding the
\enquote{best} $k$, based on a information theoretical measure of the complexity of $\mathcal{D}_k$
and the minimum description length principle~\autocite{grunwald2005tutorial}. However, in the
interest of simplicity, in the following we focus on the formulation stated in
\autoref{p:edge_full}, where $k$ is given. In practice, we further constrain $k$ to be small (say
less than \np{10}). Within the multilayer framework, $k$ could also be seen as the number of
underlying layers\stodo{elaborate on that connection}.

\begin{aside}
Generalizing the balance theory from signed graphs, we could imagine that in triangles or short
cycles, only certain combinations of directions are valid (or at least desirable). As a nice side
effect, this would likely require more innovative optimization algorithms. However, it sounds very
application-dependant. Furthermore, in our current setting, we do not know in advance in the
semantic of each direction, and we have no labeled edges. Therefore, it is not clear how those
constraints would be specified or whether they can be learned, if applicable. 
\end{aside}

Another way to leverage the graph topology is to define local constraints at the node level. Since
the beginning we considered that the user profiles were given, but from now on we further assume
that they are normalized so that they are unit norm, just like directions. In practice, if nodes are
users of a social network and the profiles measure their activity across several domains, this
allows to compare users with various level of total activity. One way to integrate that fact in
our optimization formulation is to assume that profiles are linear combination of directions, that
is $x_u = b_u + \sum_{v \in \nei(u)} a_{uv} w_{uv}$, where $a_{uv}$ are real coefficients and $b_u
\in \Rbb^d$. Viewed in the other direction, this can be interpreted as follow: each time a node $u$
connect to another node $v$ along a direction $w_{uv}$, it consume a part of its profile
proportional to $w_{uv}$. Therefore, we subtract the term $\sum_{u\in V} \left|\left| x_u - b_u -
\sum_{v \in \nei(u)} a_{uv} w_{uv} \right|\right|^2$ to \eqref{eq:edge_full}.
%
We can also make the assumption that each node $u$ is only involved in $k_\mathrm{local} < k$
different directions. The intuition is that when $d$ is large enough, users have to focus their
energy and can only express their interest in a few number of dimensions. In other words, all the
edges incident to $u$ can only be associated with one of $k_\mathrm{local}$ directions. This is
difficult to incorporate directly in the optimization formulation, as for certain value of
$k_\mathrm{local}$, this is a \NPc{} graph colouring problem. Still, it can be seen as a learning
bias and will be used to generate synthetic data.
These two constraints can be combined, so that the profiles are close to a linear combination of a
small number of directions.

@@

One can check that given a set of numbers $S=\{a_1, a_2, \ldots, a_{|S|}\}$ and real number $\beta >
0$, we have
\begin{equation*}
  \max_{a_i \in S} a_i = \lim_{\beta \rightarrow + \infty} \frac{1}{\beta}
  \log\left( \sum_{i=1}^{|S|} \exp{\beta a_i} \right)
\end{equation*}
Using this fact, and averaging the two components of our objective function, we can rewrite
\eqref{eq:edge_full} as:
\begin{equation}
  \argmax_{w_1,\ldots,w_k \in \dsphere,\, a_{uv}\in\Rbb,\, b_u \in \Rbb^d}
  \frac{1}{|E|\beta} \sum_{u,v\in E}
  \log\Bigl(\sum_{\ell=1}^k\exp\left(\beta {s_{uv}}^T w_\ell \right)\Bigr)
  - \frac{1}{|V|}\sum_{u\in V}
  \left|\left| x_u - b_u - \sum_{v \in \nei(u)} a_{uv} w_{uv} \right|\right|^2
\end{equation}

Another way to remove the $\max$ is to look for a single direction per edge, stacked horizontally in
one matrix. Then we can directly minimize \eqref{eq:edge_full} while enforcing the matrix
$W\in\mathbb{R}^{d\times |E|}$ of all directions to be low-rank, so that each $w_{uv}$ (given by the
$(u-1)n+v$-th column of $W$) can be expressed as a linear combination of a small number of basis
vectors. Formally, if $W$ has rank $k$, then there exists $P\in\mathbb{R}^{d\times k}$ and
$Q\in\mathbb{R}^{|E|\times k}$ such that $W=PQ^T$ and $w_{uv} = Pq_{uv}^T$. $P$ defines $k$
representative directions in $\Rbb^d$ while $Q$ gives, for each edge, the weights of each
representative direction. Similar to what is done in spectral clustering, we could cluster the
$q_{uv}$'s to obtain an edge labeling.
%
Denoting by $\mathbb{M}^{d\times |E|}$ the set of $d\times |E|$ matrices with unit $L_2$ norm
columns, a convex formulation of the problem is as follows:
\begin{equation}
  \label{eq:fact_convex}
  \min_{W\in\mathbb{M}^{d\times |E|}} \quad
  \sum_{i,j \in E} {s_{uv}}^T w_{uv} + \lambda \|W\|_*,
\end{equation}
where $w_{uv} \in \Rbb$ denotes the $(u-1)n+v$-th column of $W$, $\|W\|_* =
\sum_{i=1}^{\min(d,|E|)}\sigma_i(W)$ is the sum of the singular values of $W$ (called the nuclear
norm or trace norm), and $\lambda \geq 0$ is a regularization parameter.\footnote{To make the
optimal solution unique, we can use an additional Frobenius norm regularization on $W$.} The nuclear
favors low-rank solutions as it is essentially an $L_1$ norm on the spectrum.
Problem~\eqref{eq:fact_convex} can be solved using proximal gradient descent (requires singular
value decomposition at each iteration) \autocite{Parikh2013a} or Frank-Wolfe (requires only to find
the largest singular vector) \autocite{Jaggi2013a}.

Alternatively, we can consider a nonconvex formulation where we make the low-rank decomposition of
$W$ explicit. Given a rank $k\leq \min(d,|E|)$:
\begin{equation}
  \label{eq:fact_nonconvex}
  \min_{P\in\mathbb{M}^{d\times k}, Q\in\mathbb{M}^{|E|\times k}} \quad
  \sum_{u,v \in E} {s_{uv}}^T w_{uv}.
\end{equation}
Problem~\eqref{eq:fact_nonconvex} can be solved via alternating optimization over $P$ and $Q$ using
standard (projected) gradient descent algorithms.
See \autocite[Section 6]{attributedSNE17} for a write up example of matrix optimization.

Note however that this matrix formulation requires learning more parameters, as it each edge has a
different, albeit made up of a small number of basis direction.  Impose sparsity on $Q$?

Max margin, replace the max by a difference of the currently assigned direction with the other
(trade off parameter). At least I think I used it in synthetic data.

>>>>

Based on these specific definitions of $s$ and $g$, we now formally write down the problem of
finding and assigning the most informative directions to every edge of $G$
\begin{problem}[]
  \label{p:edge_full}
  Given a graph $G=(V, E)$, node profiles $X\in [-1, 1]^{n\times d}$ and an integer $k \in \Nbb$,
  find a set of $k$ unit-norm directions $\mathcal{D}_k = \{w_1, w_2, \ldots, w_k\} \in \dsphere$
  and associate to every edge of $E$ the direction with the maximal goodness. Formally, solve
  \begin{equation}
    \label{eq:edge_full}
    \argmax_{w_1,\ldots,w_k \in \dsphere}
    \sum_{u,v \in E} \max_{\ell \in \rangesk} g(s_{uv}, w_{\ell}) =
    \sum_{u,v \in E} \max_{\ell \in \rangesk} \left(x_u \circ x_v\right)^T w_{\ell} = f(k)
  \end{equation}
  where $s_{uv} = x_u \circ x_v$ and $g(s_{uv}, w_{\ell}) = {s_{uv}}^T w_{\ell}$.
\end{problem}

@@

At the other extreme of $k=2$, when $k=|E|$, the topology of the graph becomes irrelevant,
as we can simply exploit the trivial
solution of having a single direction $w_{uv} = \frac{s_{uv}}{||s_{uv}||}$ for every edge $(u,v)$.
More generally, as $k$ increases, we expect the value $\max_{\mathcal{D}_k} f(k)$ to increases as
well, at the cost of interpretability. There could be a principled approach to finding the
\enquote{best} $k$, based on a information theoretical measure of the complexity of $\mathcal{D}_k$
and the minimum description length principle~\autocite{grunwald2005tutorial}. However, in the
interest of simplicity, in the following we focus on the formulation stated in
\autoref{p:edge_full}, where $k$ is given. In practice, we further constrain $k$ to be small (say
less than \np{10}). Within the multilayer framework, $k$ could also be seen as the number of
underlying layers\stodo{elaborate on that connection}.

\stodo{move that in the conclusion}
\begin{aside}
Generalizing the balance theory from signed graphs, we could imagine that in triangles or short
cycles, only certain combinations of directions are valid (or at least desirable). As a nice side
effect, this would likely require more innovative optimization algorithms. However, it sounds very
application-dependant. Furthermore, in our current setting, we do not know in advance in the
semantic of each direction, and we have no labeled edges. Therefore, it is not clear how those
constraints would be specified or whether they can be learned, if applicable. 
\end{aside}
\stodo{Specification of constraints sounds related to network schema in \autocite[Definition
3]{HINSurvey17}, except we have a single node type and are looking at higher order patter.}

Another way to leverage the graph topology is to define local constraints at the node level. Since
the beginning we considered that the user profiles were given, but from now on we further assume
that they are normalized so that they are unit norm, just like directions. In practice, if nodes are
users of a social network and the profiles measure their activity across several domains, this
allows to compare users with various level of total activity. One way to integrate that fact in
our optimization formulation is to assume that profiles are linear combination of directions, that
is $x_u = b_u + \sum_{v \in \nei(u)} a_{uv} w_{uv}$, where $a_{uv}$ are real coefficients and $b_u
\in \Rbb^d$. Viewed in the other direction, this can be interpreted as follow: each time a node $u$
connect to another node $v$ along a direction $w_{uv}$, it consume a part of its profile
proportional to $w_{uv}$. Therefore, we subtract the term $\sum_{u\in V} \left|\left| x_u - b_u -
\sum_{v \in \nei(u)} a_{uv} w_{uv} \right|\right|^2$ to \eqref{eq:edge_full}.

We can also make the assumption that each node $u$ is only involved in $k_\mathrm{local} < k$
different directions. The intuition is that when $d$ is large enough, users have to focus their
energy and can only express their interest in a few number of dimensions. In other words, all the
edges incident to $u$ can only be associated with one of $k_\mathrm{local}$ directions. This is
difficult to incorporate directly in the optimization formulation\stodo{actually it's not that
difficult}, as for certain value of
$k_\mathrm{local}$, this is a \NPc{} graph colouring problem. Still, it can be seen as a learning
bias and will be used to generate synthetic data.
These two constraints can be combined, so that the profiles are close to a linear combination of a
small number of directions.

@@

One can check that given a set of numbers $S=\{a_1, a_2, \ldots, a_{|S|}\}$ and real number $\beta >
0$, we have
\begin{equation*}
  \max_{a_i \in S} a_i = \lim_{\beta \rightarrow + \infty} \frac{1}{\beta}
  \log\left( \sum_{i=1}^{|S|} \exp{\beta a_i} \right)
\end{equation*}
Using this fact, and averaging the two components of our objective function, we can rewrite
\eqref{eq:edge_full} as:
\begin{equation}
  \label{eq:edge_soft}
  \argmax_{{\substack{w_1,\ldots,w_k \in \dsphere,\\
  a_{uv}\in\Rbb,\, b_u \in \Rbb^d}}}
  \frac{1}{|E|\beta} \sum_{u,v\in E}
  \log\Bigl(\sum_{\ell=1}^k\exp\left(\beta {s_{uv}}^T w_\ell \right)\Bigr)
  - \frac{1}{|V|}\sum_{u\in V}
  \left|\left| x_u - b_u - \sum_{v \in \nei(u)} a_{uv} w_{uv} \right|\right|^2
\end{equation}

Another way to remove the $\max$ is to first seek one direction for every edge, that is to maximize
$\sum_{u,v\in E} {s_{uv}}^T w_{uv}$. This objective can be written in matric form as $\tr(SW) =
\frop{S^T}{W}$, where $S \in \Rbb^{|E| \times d}$ is the $s_{uv}$ vectors for all edges stacked
vertically and $W \in \Rbb^{d \times |E|}$ is the matrix of all directions stacked horizontally.
More specifically, because of the unit norm constraint on directions, we look for $W$ in the set
$\mathbb{M}^{d\times |E|}$ of $d\times |E|$ matrices with unit $\ell_2$ norm columns. To avoid the
trivial solution of letting each direction $w_{uv}$ merely be $\frac{s_{uv}}{\|s_{uv}\|}$, we
further constraint $W$ to be of low rank. That is, what we really want to optimize is \[\min_{W\in
\mathbb{M}^{d\times |E|}} -\frop{S^T}{W} + \rank(W)\,.\] However, while the Frobenius inner product is
linear in $W$, the rank operator is highly irregular and optimizing it directly is \NPc{}. Therefore
we relax the rank term by the nuclear norm of $W$, defined as $\|W\|_* = \sum_{i=1}^{\min(d,|E|)}
\sigma_i(W)$, where $\sigma_i(W)$ is the \ith{} largest singular value of $W$. The intuition is
that the rank of $W$ is
equal to the number of non-zero singular value while the nuclear norm is essentially an $\ell_1$ norm on
the spectrum of $W$. Therefore, as minimizing the $\ell_1$ enforce sparsity, minimizing the nuclear
norm tends to make the rank smaller by ensuring the presence of zeros in the singular values of $W$.
\footnote{As mentioned in \autoref{sec:troll_related} about \autocite{OnlineCompletion17}, a tighter
relaxation of the low-rank constraint is the max norm, but is it convex?} This
results in the following optimization problem: \[\min_{W\in \mathbb{M}^{d\times |E|}} -\frop{S^T}{W}
+ \lambda \|W\|_*\,,\] where $\lambda$ is a regularization parameter. Since $\|\cdot\|_*$ is a norm, this
objective function is convex (although the domain over which we optimize is not), but it does not
have a gradient, forcing us to rely on potentially costly proximal gradient descent
methods~\autocite{Parikh2013a}. Instead, we consider the equivalent problem obtained by replacing
the regularization parameter with an upper bound $\delta$ on the nuclear norm of $W$, yielding:
\begin{equation*}
  \min_{\substack{W\in \mathbb{M}^{d\times |E|} \\ \|W\|_* \leq \delta}} -\frop{S^T}{W}\,.
\end{equation*}

At this stage, we could also add the node loss term, express in matric form as $\|X - B -
C\left(A\circ W^T \right)\|_F^2$, where $B\in \Rbb^{|V| \times d}$ is the node bias vectors $b_u$
stacked vertically, $C \in \Rbb^{|V| \times |E|}$ is the incidence matrix of $G$\footnote{The
  incidence matrix of a graph $G$, usually denoted by $B$, is the $|V| \times |E|$ matrix such that
$B_{i,j} = 1$ if the node $v_i$ and edge $e_j$ are incident and 0 otherwise}, $A \in \Rbb^{|E|
\times d}$ is the matrix whose each row is the corresponding $a_{uv}\onev$ vector and $\|\cdot\|_F$
denotes the Frobenius norm. Note that this term is jointly convex in $A$, $B$ and $W$.\footnote{As
one can check that the inequality defining convexity holds for this function.} However, for the
clarity of exposition, we refrain to do so for now.

The remaining issue is that $W\in \mathbb{M}^{d\times |E|}$ is not a convex constraint. As in the
previous formulation, we thus relax it by simply imposing that the $\ell_2$ norm of the columns of
$W$ are not too large. In fact, we already have one bound on these column norms, because of the
nuclear norm constraint. One can indeed show that for $W\in \Rbb^{d\times |E|}$ such that $\|W\|_*
\leq \delta$, the \ith{} column $W_{:,i}$ of $W$ satisfies $\|W_{:,i}\|_2 \leq \delta\sqrt{d}$.
However, setting $\delta = \nicefrac{1}{\sqrt{d}}$ is too stringent in high dimension, and because
this inequality is rather loose, it will likely results in very small columns norm. Instead of
relying solely on the nuclear norm constraint, we therefore add an additional regularization term
equal to $\mu \sum_{i=1}^{|E|} \|W_{:,i}\|_2 = \mu \frop{W}{W}$. Our final convex formulation is
then:
\begin{equation}
  \label{eq:edge_FW}
  \min_{\substack{W\in \Rbb^{d\times |E|} \\ \|W\|_* \leq \delta}} \frop{\mu W - S^T}{W} \,.
\end{equation}

\begin{aside}
  Let $W$ be a matrix in $\Rbb^{d\times m}$ with $d < m$ such that $\|W\|_* \leq \delta$.
  Furthermore, let its singular value decomposition be such that we can write $W = \sum_{l=1}^d
  \sigma_l u_l v_l^T$, where $u_l$ and $v_l$ are $d$ unit-norm vectors, of dimension $d$ and $m$
  respectively. We can thus express the general term $W_{k,i}$ as $\sum_{l=1}^d \sigma_l v_{l,i}
  u_{l,k}$. Because $v_l$ is unit norm, $v_{l,i} \leq 1$ and $W_{k,i}^2 \leq \left( \sum_{l=1}^d
  \sigma_l u_{l,k} \right)^2 \leq \sum_{l=1}^d \sigma_l^2 \sum_{l=1}^d u_{l,k}^2$, using
  Cauchy--Schwartz inequality. The norm of the \ith{} column of $W$ then satisfies:
  \begin{equation*}
    \|W_{:,i}\|_2 = \left( \sum_{k=1}^m W_{k,i}^2 \right)^\lhalf
    \leq \left( \sum_{k=1}^m \left( \sum_{l=1}^d \sigma_l^2 \sum_{l=1}^d u_{l,k}^2 \right) \right)^\lhalf
    = \left( \left( \sum_{l=1}^d \sigma_l^2 \right) \left( \sum_{l=1}^d \sum_{k=1}^m u_{l,k}^2 \right) \right)^\lhalf
  \end{equation*}
  From $\sum_{l=1}^d |\sigma_l | \leq \delta$, we have that $\sum_{l=1}^d \sigma_l^2 \leq \delta^2$.
  Combined with $u_l$ being unit norm, we conclude that $\|W_{:,i}\|_2 \leq \delta\sqrt{d}$.
\end{aside}

Because \eqref{eq:edge_FW} is of the form $\min_{x\in \mathcal{X}} f(x)$ where $f$ and $\mathcal{X}$
are convex, we can solve it using the Frank--Wolfe algorithm~\autocites{FrankWolfe56}{Jaggi2013a},
which enjoy properties like a $O(\frac{1}{t})$ convergence rate, low computational cost by avoiding
projection step and sparse solution. It is an iterative procedure that maintains an estimate
$x^{(t)}$ of the solution and works succinctly as follow: it linearizes the objective $f$ at the
current position $x^{(t)}$ by computing $\nabla f(x^{(t)})$, it finds a minimum $s$ of that
linearization within the domain $\mathcal{X}$ by solving $\argmin_{s\in \mathcal{X}}
\innerp{s}{\nabla f(x^{(t)})}$ and it moves toward that minimizer by a step $\gamma$, setting
$x^{(t+1)} = (1-\gamma) x^{(t)} + \gamma s$.\footnote{$\gamma$ is either found by a line search of
set to $\nicefrac{t}{2+t}$.} In the case of \eqref{eq:edge_FW}, the linear minimization step amounts
to $\argmin_{\|W\|_* \leq \delta} 2\mu W - S^T = \argmax_{\|W\|_* \leq \delta} S^T - 2\mu W$.
Moreover, the $\delta$-ball of the nuclear norm is the convex hull of the matrix of the form $\delta
u v^T$ with $\|u\|_2 = 1 = \|v\|_2$~\autocite{Jaggi2013a}. By letting $u_1$ and $v_1$ be
respectively the left and right vectors associated with the largest singular value of $S^T - 2\mu
W$, the minimizer \enquote{$s$} we are looking for is therefore $\delta u_1 v_1^T$.

\iffalse
stacked horizontally in
one matrix. Then we can directly minimize \eqref{eq:edge_full} while enforcing the matrix
$W\in\mathbb{R}^{d\times |E|}$ of all directions to be low-rank, so that each $w_{uv}$ (given by the
$(u-1)n+v$-th column of $W$) can be expressed as a linear combination of a small number of basis
vectors. Formally, if $W$ has rank $k$, then there exists $P\in\mathbb{R}^{d\times k}$ and
$Q\in\mathbb{R}^{|E|\times k}$ such that $W=PQ^T$ and $w_{uv} = Pq_{uv}^T$. $P$ defines $k$
representative directions in $\Rbb^d$ while $Q$ gives, for each edge, the weights of each
representative direction. Similar to what is done in spectral clustering, we then cluster the
$q_{uv}$'s to obtain an edge labeling.
%
Denoting by $\mathbb{M}^{d\times |E|}$ the set of $d\times |E|$ matrices with unit $L_2$ norm
columns, a convex formulation\stodo{justify the convexity later: $L_edge$ is linear and both
$L_node$ and $\|W\|_*$ are norm and thus convex} of the problem is
as follows:
\begin{align}
  \label{eq:fact_convex}
  \max_{W\in\mathbb{M}^{d\times |E|}} h(W) &=
  \mathcal{L}_{\mathrm{edge}}^{\mathrm{mat}}(W) -
  \mu \mathcal{L}_{\mathrm{node}}^{\mathrm{mat}}(W) - \lambda \|W\|_* \\
  &= \frac{1}{|E|} \tr\left(SW\right) -
  \frac{\mu}{|V|} \|X - B - C\left(A\circ W^T \right)\|_2^2 - \lambda \|W\|_*
  % \sum_{i,j \in E} {s_{uv}}^T w_{uv} + \lambda \|W\|_*,
\end{align}
where $S \in \Rbb^{|E| \times d}$ is the $s_{uv}$ vectors for all edges stacked vertically, $\mu
\geq 0$ is a trade-off parameter, $B\in \Rbb^{|V| \times d}$ is the node bias vectors $b_u$ stacked
vertically, $C \in \Rbb^{|V| \times |E|}$ is the incidence matrix of $G$\footnote{The incidence
matrix of a graph $G$, usually denoted by $B$, is the $|V| \times |E|$ matrix such that $B_{i,j} =
1$ if the node $v_i$ and edge $e_j$ are incident and 0 otherwise}, $A \in \Rbb^{|E| \times d}$ is
the matrix whose each row is the corresponding $a_{uv}\onev$ vector, $\|\cdot\|_2$ denotes the
Frobenius norm, $\|W\|_* =
\sum_{i=1}^{\min(d,|E|)}\sigma_i(W)$ is the sum of the singular values of $W$ (called the nuclear
norm or trace norm), and $\lambda \geq 0$ is a regularization parameter.\footnote{To make the
optimal solution unique, we can use an additional Frobenius norm regularization on $W$.} The nuclear norm
favors low-rank solutions as it is essentially an $L_1$ norm on the spectrum.\footnote{As mentioned
in \autoref{sec:troll_related} about \autocite{OnlineCompletion17}, a tighter relaxation of the
low-rank constraint is the max norm.}
Both $\mathcal{L}_{\mathrm{edge}}^{\mathrm{mat}}(W)$ and
$\mathcal{L}_{\mathrm{edge}}^{\mathrm{mat}}(W)$ are convex in $W$, with respective gradient
\begin{align*}
  \frac{\partial \mathcal{L}_{\mathrm{edge}}^{\mathrm{mat}}(W)}{\partial W} &=
  \frac{1}{|E|} S^T \\
  \frac{\partial \mathcal{L}_{\mathrm{node}}^{\mathrm{mat}}(W)}{\partial W} &= \frac{-2}{|V|}
  A^T \circ \left(\Big(\left(X-B\right)^T -A^T \circ \left(W C^T\right) \Big) C\right)
\end{align*}
Problem~\eqref{eq:fact_convex} can be solved using proximal gradient descent (requires singular
value decomposition at each iteration) \autocite{Parikh2013a} or Frank-Wolfe (requires only to find
the largest singular vector) \autocite{Jaggi2013a}\stodo{look at the rest of the tutorial and
comment on the difficulty of the LMO step}.
\fi

We can avoid the rank regularization term (and thus simplify these projection issues) at the cost of
convexity by making low-rank decomposition of $W$ explicit. Given a rank $k\leq \min(d,|E|)$:
\begin{equation}
  \label{eq:fact_nonconvex}
  \max_{P\in\mathbb{M}^{d\times k}, Q\in\mathbb{M}^{|E|\times k}} \quad \tilde{h}(P, Q) =
  \mathcal{L}_{\mathrm{edge}}^{\mathrm{mat}}(PQ^T) -
  \mu \mathcal{L}_{\mathrm{node}}^{\mathrm{mat}}(PQ^T)
\end{equation}
Problem~\eqref{eq:fact_nonconvex} can be solved via alternating optimization over $P$ and $Q$ using
standard (projected) gradient descent algorithms, with the following partial derivatives:
\begin{align*}
  \frac{\partial \tilde{h}}{\partial P} &=
  \frac{1}{|E|} S^T Q + \frac{2\mu}{|V|}
  A^T \circ (((X-B)^T -A^T \circ (PQ^T )C^T )C)Q 
  \\
  \frac{\partial \tilde{h}}{\partial Q} &=
  \frac{1}{|E|} S P + \frac{2\mu}{|V|}
  A\circ (C^T (X-B-C(A\circ (QP^T ))))P.
\end{align*}
See \autocite[Section 6]{attributedSNE17} for a write up example of matrix optimization.

Compared with \refeq{eq:edge_soft}, this matrix formulation requires learning more parameters, as it
each edge has a different direction, albeit made up of a small number of basis direction.  Impose
sparsity on $Q$?

A simple and natural baseline is to cluster the set of all similarity edge vectors $
\theset{s_{uv}}{(u,v) \in E}$ using the $k$-means algorithm, where $k$ is the number of directions
set in \autoref{p:edge_full}. Formally, given the graph $G=(V,E)$ and the node profiles $X$ as
input, we order the edges from $e_1$ to $e_m$ and build the matrix $S \in \Rbb^{m\times d}$ whose
\ith{} row is the similarity between the profiles of the endpoints of $e_i=(u,v)$, that is $s_{uv} =
x_u \circ x_v$. Running $k$-means will thus partition $E$ into $k$ sets of edges, call them $E_1,
\ldots, E_k$. The first drawback of that simplicity is that it does not solve \autoref{p:edge_full}.
Indeed, the partition of $E$ does not provide the set of directions $\mathcal{D}_k = \{w_1, w_2,
\ldots, w_k\} \in \dsphere$ we are looking for. A natural way of obtaining directions from that
partition is to set $w_\ell = \frac{s_\ell}{\left\|s_\ell\right\|}$, where $s_\ell = \sum_{u,v \in
E_\ell} s_{uv}$ is the sum of all the similarity vectors of the edge in $E_\ell$. Note however that
this does not guarantee that we have assigned the best direction to every edge. We therefore propose
the following heuristic, inspired by the Lloyd algorithm for $k$-means, but where the directions in
$\mathcal{D}_k$ play the role of centroids. Namely, we alternate between two steps
\begin{enumerate}[(i), nosep]
  \item create a new partition $\{E'_1, \ldots, E'_k\}$ by assigning every edge $(u,v)$ to the
    direction $w$ maximizing its score ${s_{uv}}^T w$, that is $E'_\ell = \theset{(u,v) \in E}{
    w_\ell = \argmax_{w_i \in \mathcal{D}_k} {s_{uv}}^T w_i}$, and
  \item update the directions to make them optimal with respect to the edges currently assigned to
    them, that is as previously, set $\mathcal{D}_k = \left\{ w_\ell = \frac{s_\ell}{\left\| s_\ell
    \right\|}\right\}_{\ell=1}^k $, where $s_\ell = \sum_{u,v \in E'_\ell} s_{uv}$,
\end{enumerate}
until the edge assignment is stable or we reach a maximum number of iterations. By analogy with the
$k$-means algorithm, obtaining convergence guarantees in the general case would be surprising, but in
practice we expect that only a small number of iterations would be needed.
